---
title             : "Maternal Emotion Dysregulation and its Association with Child Internalizing and     Externalizing Behaviors and Heart Rate Variability"
shorttitle        : "Maternal Emotion Dysregulation and Child Outcomes"

author: 
  - name          : "Jackie O'Brien"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Postal address"
    email         : "my@email.com"
  - name          : "Jenn Lewis"
    affiliation   : "1"
  - name          : "Yoel Everett"
    affiliation   : "1"

affiliation:
  - id            : "1"
    institution   : "University of Oregon"

authornote: 
  



abstract: |
  Maternal emotion dysregulation, a transdiagnostic feature of psychopathology, may be a potential risk factor for the emergence of psychopathology in children. However, there is less known about child characteristics that might serve as protective factors against this risk. One such characteristic is heart rate variability (HRV) reactivity, where greater decreases in HRV from baseline to a stressor task indicate increased emotion regulation. This study examined whether increased child HRV reactivity served as a protective factor mitigating the transmission of psychopathology from emotionally dysregulated mothers to behavior problems in preschool age children.
  
  Mother-preschooler dyads (N=66) were oversampled for maternal emotion dysregulation, measured using maternal self-report on the Difficulties in Emotion Regulation Scale. Mothers reported on child internalizing and externalizing behaviors using the Child Behavioral Checklist. Child baseline HRV was collected, where the child sat quietly for 2 minutes while a book was read to them. Child HRV was also measured during a stressor task, where dyads had 7 minutes to build a complex Lego figure. HRV reactivity was calculated by subtracting child baseline HRV from child HRV during the stressor task.
  
  Two hierarchical regression models were conducted, entering maternal emotion dysregulation, child HRV reactivity, and the interaction term of these variables predicting either child internalizing or child externalizing problems (see Table 1). Across these two models, maternal emotion dysregulation, but not child HRV reactivity, significantly predicted child’s internalizing and externalizing behaviors. Maternal emotion dysregulation significantly interacted with child HRV reactivity to predict child internalizing behaviors, such that maternal emotion dysregulation had a greater impact on child internalizing behaviors if the child exhibited a greater decrease in HRV from baseline to the stressor task (i.e. exhibited increased self-regulation). There was no significant interaction predicting child externalizing behaviors. 
  
  These findings suggest that maternal emotion dysregulation more strongly predicts child behavior problems in physiologically regulated children. Interventions that target maternal emotion dysregulation may therefore improve child behavior outcomes even in physiologically regulated children.

  
keywords          : "emotion regulation, parenting, child outcomes"
wordcount         : "X"

bibliography      : ["r-references.bib"]

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup,  include=FALSE}
#You will need to install kableExtra package in order for this to run
#You will need to install jtools package in order for this to run
#Load Packages
library("papaja")
library(tidyverse)
library(rio)
library(here)
library(kableExtra)
library(jtools)
library(knitr)

knitr::opts_chunk$set()
```

```{r tidy data}
#import data and clean names
data <- import(here("data", "R Project Data copy.xlsx")) %>%
  janitor::clean_names()

#view data
#glimpse(data)


#tidy data using required functions
tidy_data <- data %>%
  filter(!is.na(family_id)) %>%
  gather(text, val, 2:3) %>%
  separate(text, c("text", "type"), sep= "_") %>%
  spread(type, val) %>%
  select(-text)

tidy_data <- data %>%
  filter(!is.na(family_id)) %>%
  gather(key = var, value = cbcl_score, starts_with("cbcl")) %>% 
  separate(var, 
           c("discard", "cbcl_subtype"), 
           sep = "_",
           convert = TRUE) %>%
  select(-discard) 

#view tidy data
#glimpse(tidy_data)

#Center DERS and Reactivity variables
center_scale <- function(x) {
  as.numeric(scale(x, scale = FALSE))
}

#Create Reactivity Variable
tidy_data <- tidy_data %>% 
  mutate(reactivity = child_lego - child_baseline, 
         ders_c = center_scale(ders), 
         reactivity_c = center_scale(reactivity))

#glimpse(tidy_data)
```

#Introduction

  Emotion dysregulation, a transdiagnostic feature of psychopathology, has been shown to be a significant mediator of mental health symptoms and symptom severity in adults. A parent's own mental health has been known to predict child mental health symptoms and behavioral problems.  These two facts together, therefore, may mean a parent's emotion regulation, particularly emotion regulation difficulties, may be an important risk factor for the emergence of psychopathology in children. Investigating the role of parental emotion regulation on childhood health and mental health problems is therfore an important clinical question in need of further investigation.

  While risk factors are one important area to investigate in the prevention of child mental health symptoms, it is also important to examine protective factors that may help make a child more resilient to developing these symptoms later on. However, there is less known about child characteristics that might serve as protective factors against risk. One such characteristic that has been identified in the literature is heart rate variability (HRV) reactivity, where greater decreases in HRV from baseline to a stressor task indicate increased emotion regulation. 
  
  This study examined whether increased child HRV reactivity served as a protective factor mitigating the transmission of psychopathology from emotionally dysregulated mothers to behavior problems in preschool age children. The aims of this research is to investigate the relationship between maternal emotion dysregulation and child behaviors in a sample of women with BPD symptoms and there preschool aged children.  A second aim is to examine the effects of maternal emotion dysregulation on child HRV reactivity. The final aim is to examine the interaction of maternal emotion dysregulation and child reactivity on child behaviors.
  
# Methods

## Participants

  Sixty-eight mothers and their preschool aged children (M = 48, SD = 7.6 months, 46% girls) were recruited from various sources including a developmental database maintained by the university psychology department, craigslist, and community mental health centers. Mothers were recruited based on the presence or absence of borderline personality disorder (BPD) symptoms, a disorder marked by extreme emotion dysregulation, as measured by the McLean screener (Zanarini et al., 2003). Mothers with elevated BPD symptoms were oversampled in order to ensure a range of emotion regulatory capabilities. 

## Procedure
Families participated in a 2.5-hour assessment in offices on a university campus. Prior to participation, both mother consent and child assent were obtained, per Institutional Review Board approval. While mothers completed questionnaires, children completed assessments in an adjacent room, although child assessment data is not presented here. Mother and children were then reunited for parent-child interaction tasks in which baseline and stressor task HRV was collected on both mothers and children. Only child HRV data is presented here. 

## Materials
**Maternal  emotion  dysregulation.** Maternal emotion dysregulation was measured using the Difficulties in Emotion Regulation Scale  (DERS; Gratz & Roemer, 2004). The DERS is a 36-item self-report questionnaire designed to assess multiple facets of emotional dysregulation, with scores ranging from 36-180 (*M*=`r mean(tidy_data$ders)`, *SD*=`r sd(tidy_data$ders)`). Higher scores suggest greater emotion dysregulation. 

**Heart rate variability.** Child baseline HRV was collected, where the child sat quietly for 2 minutes while a book was read to them. Child HRV was also measured during a stressor task, where dyads had 7 minutes to build a complex Lego figure. HRV reactivity was calculated by subtracting child baseline HRV from child HRV during the stressor task (*M*=`r mean(tidy_data$reactivity, na.rm = TRUE)`, *SD*=`r sd(tidy_data$reactivity, na.rm = TRUE)`)..

**Child behavior problems.** Child behavior problems were assessed using maternal report on the Child Behavior Checklist (CBCL) for both internalizing (i.e., anxious, depressive, and overcontrolled) and externalizing (i.e., aggressive, hyperactive, noncompliant, and undercontrolled) behaviors. Mean scores are presented in Table 2.

## Data analysis
We used `r cite_r("r-references.bib")` for all our analyses.

# Results

Means and standard deviations for variables are presented in Table 1. and Table 2.

Two hierarchical regression models were conducted, entering maternal emotion dysregulation, child HRV reactivity, and the interaction term of these variables predicting either child internalizing or child externalizing problems (see Table 2.). Across these two models, maternal emotion dysregulation, but not child HRV reactivity, significantly predicted child’s internalizing and externalizing behaviors. Maternal emotion dysregulation significantly interacted with child HRV reactivity to predict child internalizing behaviors, such that maternal emotion dysregulation had a greater impact on child internalizing behaviors if the child exhibited a greater decrease in HRV from baseline to the stressor task (i.e. exhibited increased self-regulation). There was no significant interaction predicting child externalizing behaviors. 

```{r descriptives ders and reactivity, results='asis'}
#str(tidy_data)

#Descriptives Table DERS and Reactivity
table_ders_reactivity <- tidy_data %>% 
  summarize(DERS_mean = mean(ders, na.rm = TRUE), 
            DERS_SD = sd(ders, na.rm = TRUE), 
            Reactivity_mean = mean(reactivity, na.rm = TRUE), 
            Reactivity_SD = sd(reactivity, na.rm = TRUE))

table_ders_reactivity[, -1] <-printnum(table_ders_reactivity[, -1])

apa_table(table_ders_reactivity,
          caption = "Means and SDs for Maternal Emotion Dysregulation (DERS) and Child Reactivity")

#kable(table, 
#      format = "latex",
#      booktabs = TRUE,
#      caption = "Means and SDs of maternal emotion dysregulation (DERS) and child reactivity")

```

```{r descriptives cbcl, results='asis'}
#Descriptives Table Internalizing and Externalizing Behavior
table_cbcl <- tidy_data %>%
  group_by(cbcl_subtype) %>%
  summarize(cbcl_mean= round(mean(cbcl_score, na.rm = TRUE), 2), 
            cbcl_SD = sd(cbcl_score, na.rm = TRUE))

table_cbcl[, -1] <-printnum(table_cbcl[, -1])

apa_table(table_cbcl,
          caption = "Means and SDs for Child Internalizing and Externalizing Behavior")


#kable(table2, 
 #     format = "latex",
  #    booktabs = TRUE,
  #    caption = "Means and SDs of child internalizing and externalizing behavior")

```

```{r linear regression internalizing, echo=FALSE, results='asis'}
linear_model1 <- lm(cbcl_score ~ ders_c * reactivity_c, 
                    data = subset(tidy_data, cbcl_subtype == "int"))
#summary(linear_model1)

linear_model1_print <- apa_print(linear_model1)

apa_table(linear_model1_print$table,
          caption = "Results of Linear Regression Predicting Child Internalizing Behavior")

#Tried this code from sjPlot package, but it will not knit to pdf
#tab_model(
# linear_model1,
#  pred.labels = c("Intercept", "Maternal Em. Dysreg", "Child Reactivity", "Interaction"),
#  dv.labels = c("Internalizing Behavior (CBCL)"),
#  string.pred = "Coefficient",
#  string.ci = "CI (95%)",
#  string.p = "P-Value"
#  )
```

```{r linear regression externalizing, echo=FALSE, results='asis'}
linear_model2 <- lm(cbcl_score ~ ders_c * reactivity_c, 
                    data = subset(tidy_data, cbcl_subtype == "ext"))

linear_model2_print <- apa_print(linear_model2)

apa_table(linear_model2_print$table,
          caption = "Results of Linear Regression Predicting Child Externalizing Behavior")

#summary(linear_model2)

#stargazer(linear_model2, type = "text", title = "Linear Model 2", digits = 2)

#library(sjPlot)
#tab_model(
#  linear_model2,
#  pred.labels = c("Intercept", "Maternal Em. Dysreg", "Child Reactivity", "Interaction"),
#  dv.labels = c("Externalizing Behavior (CBCL)"),
#  string.pred = "Coefficient",
#  string.ci = "CI (95%)",
#  string.p = "P-Value"
#  )
```


```{r plot1, fig.cap="Maternal Emotion Dysregulation and Child Behaviors", warning=FALSE}
#A scatterplot of maternal emotion dysregulation (x-axis)
#and child behaviors (y-axis), including a regression line 
ggplot(tidy_data, aes(x=ders, y=cbcl_score, color = cbcl_subtype)) +
  geom_point(color="black") +
  geom_smooth(method = "lm") +
   labs(x="Maternal Emotion Dysregulation (DERS)",
       y="Child Problem Behaviors")
```

```{r plot2, fig.cap="Maternal Emotion Dysregulation and Child HRV Reactivity", warning=FALSE}
#A scatterplot of maternal emotion dysregulation (x-axis)
#and reactivity (y-axis), including a regression line 
ggplot(tidy_data, aes(x=ders, y=reactivity)) +
  geom_point(color="black") +
  geom_smooth(method = "lm") +
   labs(x="Maternal Emotion Dysregulation (DERS)",
       y="Child Reactivity HRV")
```

```{r plot3, fig.cap= "Child Reactivity Predicting Child Behavior Problems at Three Different Levels of Maternal Emotion Dysregulation "}
interaction_plot <- interact_plot(linear_model1, pred = "reactivity_c", modx = "ders_c", centered = "none", 
      x.label="Child Reactivity HRV",
      y.label="Child Behavior Problems (CBCL)",
      legend.main = "DERS")
interaction_plot
```



# Discussion
These findings suggest that maternal emotion dysregulation more strongly predicts child behavior problems in physiologically regulated children. Interventions that target maternal emotion dysregulation may therefore improve child behavior outcomes even in physiologically regulated children. 

\newpage

# References
```{r create_r-references}
r_refs(file = "r-references.bib")
```

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id = "refs"></div>
\endgroup
